Evaluating infectious disease forecasts with allocation scoring rules

### 2023-12-01 alloscore_manuscript/alloscore-application.Rnw
manuscript and figure updates

To illustrate the mechanics of allocation scoring, we start by focusing on how forecasts generated on or before December 20, 2021, with predictions for January 3, 2022, were scored by different metrics.
This week was around the Omicon wave peak nationally, with individual states typically observing a peak at or after January 3, 2022.

For many locations, forecasts were \ngr{fill in details here}.
On this date, of the 10 models evaluated, the CU-select model had the most accurate forecasts according to the allocation score while the USC-SI\_kJalpha model had the most accurate forecasts based on MWIS (Table \ref{tab:one-week-results}).
The JHUAPL-Bucky model had the second best MWIS but the third worst allocation score.

<<metrics-rankings, results = "asis">>=
## for highlighting cells, if we want that https://stackoverflow.com/questions/37448348/color-code-cells-of-xtable
all_scores |>
  filter(target_end_date == "2022-01-10", !is.na(mwis)) |>
  select(model, alloscore, allo_rank, mwis, mwis_rank) |>
  arrange(allo_rank) |>
  xtable::xtable(label = "tab:one-week-results",
                 caption = "caption TBD") |>
  print(include.rownames = FALSE)
@

Forecasts and allocations for the nine states with the highest numbers of hospitalizations on January 3, 2022 reveal mechanics about how allocations are made (Figure \ref{fig:thermometer-plot}).
As described in the methods above, allocations for a given location from a specific model are generated by finding the single quantile of the probabilistic forecast across all locations that reaches the resource allocation limit (in this case, assumed to be 15,000).
A direct comparison between the forecasts from the JHUAPL-Bucky and USC-SI\_kJalpha models show that while the JHUAPL-Bucky forecast distributions were closer to the truth across all states, the allocations suggested by those forecasts often were more inefficient than those from the USC-SI\_kJalpha model (Figure \ref{fig:thermometer-plot}).
\ngr{add in some specific interpretations from the figure.}

@@

<<multi-loc-ranks>>=
## date points to target_end_date of 2022-01-10 to highlight
ref_date_for_plot <- "2021-12-20"
models_to_plot <- c("JHUAPL-Bucky", "CU-select")

# tar_load(forecast_data)
# forecast_dat_to_merge <- forecast_data |>
#   dplyr::filter(reference_date == ref_date_for_plot,
#          model %in% c("JHUAPL-Bucky", "MUNI-ARIMA"),
#          quantile == "0.5") |>
#   select(model, reference_date, state = abbreviation, quantile, median = value) |>
#   mutate(reference_date = as.character(reference_date))

multiloc_dat <- Kat15k_alloscores |>
  filter(reference_date == ref_date_for_plot,
         ## the above date points to target_end_dateof 2022-01-10 where things were really wrong
         model %in% models_to_plot) |>
  rename(allocation = x, observation = y, state = abbreviation) |>
  select(model, reference_date, state, allocation, observation) |>
#  left_join(forecast_dat_to_merge) |>
#  pivot_longer(cols = c("allocation", "observation", "median"))
  pivot_longer(cols = c("allocation", "observation"))

## plot all locations observations and allocations
multiloc_plot <- ggplot(multiloc_dat, aes(x=name, y=value, group=state)) +
  geom_line() +
  geom_line(data=dplyr::filter(multiloc_dat, state=="FL"), color="orange") +
  ylab("Daily hospital admissions") +
  scale_x_discrete(name=NULL, limits=rev, expand = c(.1,.1)) +
  facet_grid(.~model)

oracle_rows <- Kat15k_alloscores |>
  filter(reference_date == ref_date_for_plot,
         model %in% c("JHUAPL-Bucky")) |>
  rename(observation = y, state = abbreviation, allocation = oracle) |>
  select(reference_date, state, allocation, observation) |>
  pivot_longer(cols = c("allocation", "observation")) |>
  mutate(model = "Oracle")

multiloc_dat |>
  bind_rows(oracle_rows) |>
  mutate(state = reorder(state, -ifelse(name =="observation", value, 0)),
         name = factor(name, levels = c("observation", "allocation"))) |>
  arrange(name) |>
  ggplot(aes(x=state, y=value, fill=name)) +
  scale_fill_manual(values = c("#CD2626", "goldenrod"))+
  geom_col(position="identity", alpha=.7) +
  facet_grid(model~.)
@

>>>>

To illustrate the mechanics of allocation scoring, we start by focusing on how forecasts generated on or before December 20, 2021, with predictions for January 03, 2022, were scored by different metrics.
This week was around the Omicon wave peak nationally, with individual states typically observing a peak at or after January 3, 2022.

For many locations, forecasts predicted lower values than were eventually observed, as this was during the period of steep increase of viral transmission across many states.
Of the 10 models evaluated, the CU-select model had the most accurate forecasts according to the allocation score while the USC-SI\_kJalpha model had the most accurate forecasts based on MWIS (Table \ref{tab:one-week-results}).
The JHUAPL-Bucky model had the second best MWIS but the third worst allocation score.

<<metrics-rankings, results = "asis">>=
## for highlighting cells, if we want that https://stackoverflow.com/questions/37448348/color-code-cells-of-xtable
all_scores |>
  filter(target_end_date == "2022-01-03", !is.na(mwis)) |>
  select(model, alloscore, allo_rank, mwis, mwis_rank) |>
  arrange(allo_rank) |>
  xtable::xtable(label = "tab:one-week-results",
                 caption = "caption TBD") |>
  print(include.rownames = FALSE)
@

Forecasts and allocations for a selection of states with high numbers of hospitalizations on January 3, 2022 reveal mechanics about how allocations are made (Figure \ref{fig:thermometer-plot}).
As described in the methods above, allocations for a given location from a specific model are generated by finding the single quantile of the probabilistic forecast across all locations that reaches the resource allocation limit (in this case, assumed to be 15,000).
A direct comparison between the forecasts from the JHUAPL-Bucky and CU-select models show that while the JHUAPL-Bucky forecast distributions were closer to the eventual observations in many states, the allocations suggested by those forecasts often were more inefficient than those from the CU-select model (Figure \ref{fig:thermometer-plot}).
For JHUAPL-Bucky, the allocations were assigned around the 25th quantile of the predictive distribution and for CU-select it was around the 90th percentile. \ngr{could we make these more specific?}
This reflects that in general the JHUAPL-Bucky forecasts assigned more probabilities to higher values and thus a low quantile value across all locations reached the allocation limit of 15,000.
In California, New York and Texas, the eventually observed number of hospital admissions was closer to the predicted median from JHUAPL-Bucky and was contained in JHUAPL-Bucky's 80\% PI. However for each of these three states, the resource allocation was lower and therefore resulted in more hypothetical unmet need than CU-select allocations.
In Florida, the forecast from JHUAPL-Bucky missed by a wide margin (as did the forecast from CU-select) and the allocation was nearly XX units lower than what was allocated according to the CU-select model. \ngr{add specific number here}

@@

In the example of this one week, JHUAPL-Bucky had a worse allocation score than CU-select because its forecasts led to allocations that sent excess resources to several states, such as Ohio, Pennsylvania, and Michigan, that would have been more effectively allocated to states that did receive enough resources to meet their needs, such as Florida and California (Figure \ref{multi-loc-ranks}A). These allocation errors resulted because that model’s forecasts did not consistently capture the relative resource needs across different states. (Figure \ref{fig:thermometer-plot} \& \ref{multi-loc-ranks}A). The CU-select model made some similar errors — most prominently, over-allocating resources to Ohio — but overall, it did a better job of forecasting the relative resource demands across different locations.

On the other hand, CU-select had worse performance as measured by WIS. Its forecasts were biased downwards, and it consistently incurred a large penalty for underprediction (Figure \ref{fig:multi-loc-rank}B). Predictions from the JHUAPL-Bucky model were wider, and included the observed level of daily hospital admissions more often. Therefore, WIS did not as often assign that model severe penalties for forecasts that underpredicted or overpredicted the actual hospitalization count.

Together, these results corroborate the understanding that the allocation score and weighted interval score measure different aspects of forecast performance. The allocation score penalizes models that mischaracterize the relative magnitude of resource need across different locations, while WIS penalizes forecasts that do not capture the absolute magnitude of the target quantity in each location.

<<multi-loc-ranks, fig.cap="Component-wise breakdowns of the allocation score (Panel A) and weighted interval score (Panel B), by location for forecasts of hospitalization admissions on January 3, 2022, for two selected models (JHUAPL-Bucky and CU-select). Panel A shows the observed resource need, in this case the observed number of hospitalizations, for each state, along with the hypothetical number of resources allocated to the given location based on the forecasts from each model. The number of available resources was fixed at 15,000 and forecasts from each model were used to determine an optimal allocation strategy before the resource need was known. For most locations the resource need exceeded the resources allocated, indicated by the 'observed resource need' bar being larger than the 'allocation' bar. Panel B shows the breakdown of the weighted interval score (WIS) into components of underprediction, overprediction and dispersion. Larger values of WIS indicate more error, and the full WIS score for each location can be decomposed into the three components shown here.">>=
## date points to target_end_date of 2022-01-10 to highlight
ref_date_for_plot <- "2021-12-20"
models_to_plot <- c("JHUAPL-Bucky", "CU-select")

multiloc_dat <- Kat15k_alloscores |>
  filter(reference_date == ref_date_for_plot,
         ## the above date points to target_end_date of 2022-01-03 where things were really wrong
         model %in% models_to_plot) |>
  rename(allocation = x, `observed resource need` = y, state = abbreviation) |>
  select(model, reference_date, state, allocation, `observed resource need`) |>
  pivot_longer(cols = c("allocation", "observed resource need"))

## plot all locations observations and allocations

## line plot below here
# multiloc_plot <- ggplot(multiloc_dat, aes(x=name, y=value, group=state)) +
#   geom_line() +
#   geom_line(data=dplyr::filter(multiloc_dat, state=="FL"), color="orange") +
#   ylab("Daily hospital admissions") +
#   scale_x_discrete(name=NULL, limits=rev, expand = c(.1,.1)) +
#   facet_grid(.~model)

ordered_levels <- truth_data |>
  filter(target_end_date == as.Date(ref_date_for_plot)+14, location != "US") |>
  select(location, abbreviation, value) |>
  arrange(-value)

alloc_plot <- multiloc_dat |>
  mutate(state = factor(state, levels = ordered_levels$abbreviation),
         name = factor(name, levels = c("observed resource need", "allocation"))) |>
  arrange(name) |>
  ggplot(aes(x=state, y=value, fill=name)) +
  scale_fill_manual(name=NULL, values = c("#CD2626", "goldenrod")) +
  theme(axis.text.x=element_text(angle=90,vjust=.5), legend.position = c(.9,.95), legend.justification = c(1,1)) +
  ylab("Resource need and allocation") +
  xlab(NULL) +
  geom_col(position="identity", alpha=.7) +
  facet_grid(model~.)

wis <- covidHubUtils::score_forecasts(
  forecast_data |> filter(reference_date == "2021-12-20", str_detect(model,"Buck|CU-select")),
  truth_data,
  use_median_as_point = TRUE,
  metrics = "wis_components",
  return_format="long") |>
  filter(location != "US", score_name != "true_value") |>
  left_join(select(ordered_levels, location, abbreviation)) |>
  mutate(abbreviation = factor(abbreviation, levels=ordered_levels$abbreviation),
         `WIS component` = factor(score_name, levels = c("overprediction", "dispersion", "underprediction")))

## palette from
## colorspace::divergingx_hcl(palette = "Blue-Yellow 3", n=3)

wis_plot <- ggplot(data = wis, aes(x=abbreviation)) +
  geom_col(aes(y=score_value, fill=`WIS component`)) +
  scale_fill_manual(values = c("#9FA2FF", "gray77", "#BAAE00")) +
  facet_grid(model~.) +
  theme(axis.text.x=element_text(angle=90,vjust=.5), legend.position = c(.9,.95), legend.justification = c(1,1)) +
  xlab(NULL) +
  ylab("WIS")

cowplot::plot_grid(alloc_plot, wis_plot,
                   nrow=2,
                   align="v",
                   rel_heights = c(1, 1),
                   labels = "AUTO")
